% !TEX root = twibop2.tex

\chapter*{Recommended Literature}
\addcontentsline{toc}{chapter}{Recommended Literature}

{\small\parindent=0pt
\section*{Chapter 1}

Chubarev, A.M., and Kholodny V.S., \emph{The Improbable Probability}, Znanie, Moscow, 1976 (in Russian).

Khurgin, Ya.L., \emph{How to Embrace the Unembraceable}, Znanie, Moscow, 1979 (in Russian).

R\'enyi, A., \emph{Levelek a val\'oszin\"us\'egr\"ol}, Akademiai kiado, Budapest, 1969.

Wentzel, E.S., \emph{Probability Theory (first steps)}, Mir Publishers, Moscow, 1982.

\section*{Chapter 2}

Khurgin, Va.L., \emph{Yes, No. or May Be \ldots}, Mir Publishers, Moscow, 1985.

Platonov, G.A., Fainberg, M. A., and Shtil'man, M. S., \emph{Trains, Passengers, and Mathematics}, Transport, Moscow, 1977 (in Russian).

Rastrigin, L., \emph{This Chancy, Chancy, Chancy World}, Mir Publishers, Moscow, 1984.

Venttsel, Ye.S., (E. S. Wentzel), \emph{Elements of Game Theory}, Mir Publishers, Moscow, 1984.

Wentzel, E.S., \emph{Operations Research. A Methodological Approach}, Mir Publishers, Moscow, 1983.

\section*{Chapter 3}

Artobolevsky, I.I., and Kobrinsky, A.E., \emph{Meet Robots}, Molodaya Gvardiya, Moscow, 1979 (in Russian).

Pekelis, V., \emph{Cybernetics A to Z}, Mir Publishers, Moscow, 1974.

Rastrigin, L., \emph{This Chancy, Chancy, Chancy World}, Mir Publishers, Moscow, 1984.

R\'enyi, A., \emph{Napl\'o az inform\'acioelm\'eletr\'ol}, Gondolat, Budapest, 1976.

Rosenblutt, M. (ed.), \emph{Studies in Probability Theory} (MAA Studies in Mathematics), Vol. 18, John Wiley \& Sons, Ltd., 1980.

Teplov, L.P., \emph{Essays on Cybernetics}, Moskovskii Rabochii, Moscow, 1963 (in Russian).

\section*{Chapter 4}

Chambadal, P., \emph{Evolution et applications du concept d'entropie}, Dunod, Paris, 1963.

Kompaneets, A.S., \emph{The Laws of Physical Statistics}, Nauka, Moscow, 1976 (in Russian).

Smorodinsky, Ya.A., \emph{Temperature}, Mir Publishers, Moscow, 1984.

\section*{Chapter 5}


Batygin, V.V., \emph{The Laws of the Microcosm}, Prosveshchenie, Moscow, 1981 (in Russian).

K\'arolyh\'azy, F., \emph{Igaz var\'azslat}, Gondolat, Budapest, 1976.

Kompaneets, A.S., \emph{What is Quantum Mechanics?} Nauka, Moscow, 1977 (in Russian).

Ponomarev, L.I., \emph{In Quest of the Quantum}, Mir Publishers, Moscow, 1973.

\section*{Chapter 6}

Luchnik, N., \emph{Why I'm Like Dad?} Mir Publishers, Moscow, 1978.

Mednikov, B.M., \emph{The Axioms of Biology}, Znanie, Moscow, 1982 (in Russian).

Polynin, V.M., \emph{Mother, Father, and Me}, Sovetskaya Rossiya, Moscow, 1967 (in Russian).

Reznik, S.E., \emph{Nikolai Vavilov}, Molodaya Gvardiya, Moscow, 1968 (in Russian).

Tarasov, L., \emph{This Amazingly Symmetrical World}, Mir Publishers, Moscow, 1987.
}
%%% Local Variables:
%%% mode: latex
%%% TeX-engine: xetex
%%% TeX-master: "twibop2"
%%% End:
